---
title: "Spatial Regression Models"
output:
  rmarkdown::html_vignette:
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
vignette: >
  %\VignetteIndexEntry{Spatial Regression Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
  - \def\T{{ \scriptstyle \top }}
  - \newcommand{\thetasp}{{\theta_{\text{sp}}}}
  - \newcommand{\GP}{\mathsf{GP}}
  - \newcommand{\N}{\mathsf{N}}
  - \newcommand{\EF}{\mathsf{EF}}
  - \newcommand{\Norm}{\mathsf{N}}
  - \newcommand{\GCMc}{\mathsf{GCM}_c}
  - \newcommand{\GCM}{\mathsf{GCM}}
  - \newcommand{\calL}{\mathcal{L}}
  - \newcommand{\IG}{\mathsf{IG}}
  - \newcommand{\IW}{\mathsf{IW}}
  - \newcommand{\given}{\mid}
bibliography: refs.bib
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this article, we discuss the following functions -

- `spLMexact()`
- `spLMstack()`
- `spGLMexact()`
- `spGLMstack()`
- `stvcLMexact()` and `stvcLMstack()` (spatially varying coefficients)

These functions can be used to fit Gaussian and non-Gaussian spatial point-referenced data.

```{r}
set.seed(1729)
```

## Bayesian Gaussian spatial regression models

In this section, we thoroughly illustrate our method on synthetic Gaussian as well as non-Gaussian spatial data and provide code to analyze the output of our functions. We start by loading the package.
```{r setup}
library(spStack)
library(ggplot2)
library(patchwork)
```
The package lazy-loads the synthetic dataset `simSpatial`, observed at 500 locations in the unit square. It has two covariates `x1` and `x2`, a spatial effect `z_true` with a ripple pattern, $z(s) = 1.5 \sin(4\pi \lVert s - (0.5, 0.5) \rVert)$, and one response for each family: Gaussian `y_gauss`, Poisson `y_pois`, binomial `y_binom` (out of `n_trials` trials) and binary `y_bin`. All the responses share the covariates and the spatial effect, so we apply every function in this article to the same data. The spatial effect is a deterministic surface, not a draw from a Gaussian process, so we can see how well the Gaussian process models recover it. See `?simSpatial` for the code that generated the data.

```{r fig.align='center', fig.alt="Interpolated surface of the true spatial effect."}
data("simSpatial")
surfaceplot(simSpatial, coords_name = c("s1", "s2"), var_name = "z_true")
```

### Using fixed hyperparameters
We first take the first 200 rows of the data and set up the priors. Supplying the priors is optional. See the documentation of `spLMexact()` to learn more about the default priors. Besides, setting the priors, we also fix the values of the spatial process parameters (spatial decay $\phi$ and smoothness $\nu$) and the noise-to-spatial variance ratio ($\delta^2$).
```{r}
dat <- simSpatial[1:200, ] # work with first 200 rows

muBeta <- c(0, 0, 0)
VBeta <- diag(1E4, 3)
sigmaSqIGa <- 2
sigmaSqIGb <- 2
phi0 <- 6
nu0 <- 0.5
noise_sp_ratio <- 0.8
prior_list <- list(beta.norm = list(muBeta, VBeta),
                   sigma.sq.ig = c(sigmaSqIGa, sigmaSqIGb))
nSamples <- 1000
```

We define the spatial model using a `formula`, similar to that in the widely used `lm()` function in the `stats` package. Here, the formula `y_gauss ~ x1 + x2` corresponds to the spatial linear model $$y(s) = \beta_0 + \beta_1 x_1(s) + \beta_2 x_2(s) + z(s) + \epsilon(s)\;,$$ where `y_gauss` corresponds to the response variable $y(s)$, which is regressed on the predictors `x1` and `x2` given by $x_1(s)$ and $x_2(s)$. The intercept is automatically considered within the model, and hence `y_gauss ~ x1 + x2` is functionally equivalent to `y_gauss ~ 1 + x1 + x2`. Moreover, a spatial random effect is inherent in the model, where the spatial correlation matrix is governed by the spatial correlation function specified by the argument `cor.fn`. Supported correlation functions are `"exponential"` and `"matern"`. The exponential covariogram is specified by the hyperparameter $\phi$ and the Matern covariogram is specified by the hyperparameters $\phi$ and $\nu$. Fixed values of these hyperparameters are supplied through the argument `spParams`. In addition, the noise-to-spatial variance ration is also fixed through the argument `noise_sp_ratio`.

If interested in calculation of leave-one-out predictive densities (LOO-PD), `loopd`  must be set `TRUE` (the default is `FALSE`). Method of LOO-PD calculation can be also set by the option `loopd.method` which support the keywords `"exact"` and `"psis"`. The option `"exact"` exploits the analytically available expressions of the predictive density and implements an efficient row-deletion Cholesky factor update for fast calculation and avoids refitting the model $n$ times, where $n$ is the sample size. On the other hand, `"psis"` implements Pareto-smoothed importance sampling and finds approximate LOO-PD and is much faster than `"exact"`.

We pass these arguments into the function `spLMexact()`.
```{r spLMexactLOO_exact}
mod1 <- spLMexact(y_gauss ~ x1 + x2, data = dat,
                  coords = as.matrix(dat[, c("s1", "s2")]),
                  cor.fn = "matern",
                  priors = prior_list,
                  spParams = list(phi = phi0, nu = nu0),
                  noise_sp_ratio = noise_sp_ratio, n.samples = nSamples,
                  loopd = TRUE, loopd.method = "exact",
                  verbose = TRUE)
```

Next, we can summarize the posterior samples of the fixed effects as follows.
```{r}
post_beta <- mod1$samples$beta
summary_beta <- t(apply(post_beta, 1, function(x) quantile(x, c(0.025, 0.5, 0.975))))
rownames(summary_beta) <- mod1$X.names
print(summary_beta)
```

### Leave-one-out predictive densities using PSIS
Out of curiosity, we find the LOO-PD for the same model using the approximate method that uses Pareto-smoothed importance sampling, or PSIS. See @LOOCV_vehtari17 for details.
```{r spLMexactLOO_PSIS}
mod2 <- spLMexact(y_gauss ~ x1 + x2, data = dat,
                  coords = as.matrix(dat[, c("s1", "s2")]),
                  cor.fn = "matern",
                  priors = prior_list,
                  spParams = list(phi = phi0, nu = nu0),
                  noise_sp_ratio = noise_sp_ratio, n.samples = nSamples,
                  loopd = TRUE, loopd.method = "PSIS",
                  verbose = FALSE)
```

Subsquently, we compare the LOO-PD obtained by the two methods.
```{r fig.align='center'}
loopd_exact <- mod1$loopd
loopd_psis <- mod2$loopd
loopd_df <- data.frame(exact = loopd_exact, psis = loopd_psis)

library(ggplot2)
plot1 <- ggplot(data = loopd_df, aes(x = exact)) +
  geom_point(aes(y = psis), size = 1, alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", alpha = 0.5) +
  xlab("Exact") + ylab("PSIS") + theme_bw() +
  theme(panel.background = element_blank(),
        panel.grid = element_blank(), aspect.ratio = 1)
plot1
```

### Using predictive stacking
Next, we move on to the Bayesian spatial stacking algorithm for Gaussian data. We supply the same prior list and provide candidate models constructed using `candidateModels()`.
```{r spLMstack}
cand.mod <- candidateModels(list(phi = c(3, 6, 10),
                                 nu = c(0.5, 1, 1.5),
                                 noise_sp_ratio = c(0.5, 1.5)),
                            "cartesian")

mod3 <- spLMstack(y_gauss ~ x1 + x2, data = dat,
                  coords = as.matrix(dat[, c("s1", "s2")]),
                  cor.fn = "matern",
                  priors = prior_list,
                  candidate.models = cand.mod,
                  n.samples = 1000, loopd.method = "exact",
                  parallel = FALSE, verbose = TRUE)
```

The user can check the solver used for the stacking weights, its status, and the runtime by issuing the following. The `diagnostics` element also holds numerical diagnostics of each candidate model.
```{r}
print(mod3$diagnostics$solver$used)
print(mod3$diagnostics$solver$status)
print(mod3$run.time)
```

### Analyzing samples from the stacked posterior
To sample from the stacked posterior, the package provides a helper function called `stackedSampler()`. Subsequent inference proceeds from these samples obtained from the stacked posterior.
```{r}
post_samps <- stackedSampler(mod3)
```

We then collect the samples of the fixed effects and summarize them as follows.
```{r}
post_beta <- post_samps$beta
summary_beta <- t(apply(post_beta, 1, function(x) quantile(x, c(0.025, 0.5, 0.975))))
rownames(summary_beta) <- mod3$X.names
print(summary_beta)
```

The response `y_gauss` was simulated using the true value $\beta = (2, 5, -1)^\T$. We notice that the stacked posterior is concentrated around the truth.
```{r, fig.align='center', fig.height=3, fig.width=8, fig.alt="Posterior distributions of the fixed effects"}
library(tidyr)
library(dplyr)

post_beta_df <- as.data.frame(post_beta)
post_beta_df <- post_beta_df %>%
  mutate(row = paste0("beta", row_number()-1)) %>%
  pivot_longer(-row, names_to = "sample", values_to = "value")

# True values of beta0, beta1 and beta2
truth <- data.frame(row = c("beta0", "beta1", "beta2"), true_value = c(2, 5, -1))

ggplot(post_beta_df, aes(x = value)) +
  geom_density(fill = "lightblue", alpha = 0.6) +
  geom_vline(data = truth, aes(xintercept = true_value),
             color = "red", linetype = "dashed", linewidth = 0.5) +
  facet_wrap(~ row, scales = "free") + labs(x = "", y = "Density") +
  theme_bw() + theme(panel.background = element_blank(),
                     panel.grid = element_blank(), aspect.ratio = 1)
```

Furthermore, we compare the posterior samples of the spatial random effects with their corresponding true values.
```{r fig.align='center', fig.alt="Comparison of stacked posterior with the true values"}
post_z <- post_samps$z
post_z_summ <- t(apply(post_z, 1, function(x) quantile(x, c(0.025, 0.5, 0.975))))
z_combn <- data.frame(z = dat$z_true, zL = post_z_summ[, 1],
                      zM = post_z_summ[, 2], zU = post_z_summ[, 3])

plotz <- ggplot(data = z_combn, aes(x = z)) +
  geom_point(aes(y = zM), size = 0.75, color = "darkblue", alpha = 0.5) +
  geom_errorbar(aes(ymin = zL, ymax = zU), width = 0.05, alpha = 0.15,
                color = "skyblue") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  xlab("True z") + ylab("Stacked posterior of z") + theme_bw() +
  theme(panel.background = element_blank(),
        panel.grid = element_blank(), aspect.ratio = 1)
plotz
```

The package also provides helper functions to plot interpolated spatial surfaces in order for visualization purposes. The function `surfaceplot()` creates a single spatial surface plot, while `surfaceplot2()` creates two side-by-side surface plots. We are using the later to visually inspect the interpolated spatial surfaces of the true spatial effects and their posterior medians.
```{r fig.align='center', fig.height=3.5, fig.width=7, fig.alt="Comparison of the interploated spatial surfaces of the true random effects and the posterior medians."}
postmedian_z <- apply(post_z, 1, median)
dat$z_hat <- postmedian_z
plot_z <- surfaceplot2(dat, coords_name = c("s1", "s2"),
                       var1_name = "z_true", var2_name = "z_hat")
patchwork::wrap_plots(plot_z) +
    patchwork::plot_layout(guides = "collect") &
    ggplot2::theme(legend.position = "right")
```

### Spatially varying coefficients
The effect of a covariate may itself vary over space. The functions `stvcLMexact()` and `stvcLMstack()` fit the Gaussian linear model with spatially-temporally varying coefficients
$$y(\ell) = x(\ell)^\T \beta + \tilde{x}(\ell)^\T z(\ell) + \epsilon(\ell), \quad \epsilon(\ell) \sim \N(0, \sigma^2),$$
where each of the $r$ varying coefficients $z_j$ is an independent Gaussian process with variance $\sigma^2_{z_j} = \sigma^2/\delta^2_j$ and a Gneiting space-time correlation function with decay parameters $\phi_s$ and $\phi_t$. The joint posterior is available in closed form and is sampled exactly. If all observations share one time point, the Gneiting correlation reduces to the exponential correlation $\exp(-\phi_s \lVert s - s' \rVert)$ ($\phi_t$ then has no effect), and these functions fit a spatially varying coefficients model.

The response `y_svc` of `simSpatial` was simulated from $$y(s) = 2 + \{5 + z(s)\} x_1(s) - x_2(s) + \epsilon(s), \quad \epsilon(s) \sim \N(0, 0.5^2)\;,$$ so that the ripple $z(s)$ is now the spatially varying part of the slope of $x_1$. In the `formula`, the variables in parentheses receive varying coefficients: `y_svc ~ x1 + x2 + (x1)` has fixed effects for the intercept, `x1` and `x2`, and varying coefficients for the intercept and `x1` (the intercept is included in both parts by default). With `process.type = "independent.shared"`, the processes share the decay parameters and the noise-to-spatial variance ratio; with `process.type = "independent"`, each process has its own, supplied as vectors of length $r$. We supply a constant time coordinate.
```{r stvcLMexact_svc}
n <- nrow(dat)
mod4 <- stvcLMexact(y_svc ~ x1 + x2 + (x1), data = dat,
                    sp_coords = as.matrix(dat[, c("s1", "s2")]),
                    time_coords = matrix(0, n, 1),
                    cor.fn = "gneiting-decay",
                    process.type = "independent.shared",
                    sptParams = list(phi_s = 3, phi_t = 1),
                    noise_sp_ratio = 0.2,
                    n.samples = 1000, loopd = TRUE, verbose = TRUE)
```

The posterior samples of the fixed effects, the noise variance $\sigma^2$, and the process variance $\sigma^2_z = \sigma^2/\delta^2$ are summarized below.
```{r}
summary_beta <- t(apply(mod4$samples$beta, 1, function(x) quantile(x, c(0.025, 0.5, 0.975))))
rownames(summary_beta) <- mod4$X.names
print(summary_beta)
print(quantile(mod4$samples$sigmaSq, c(0.025, 0.5, 0.975)))
print(quantile(mod4$samples$sigmaSq.z, c(0.025, 0.5, 0.975)))
```

Next, we use predictive stacking over candidate values of $\phi_s$ and $\delta^2$. For `process.type = "independent"`, each candidate value is a vector of length $r$, supplied as `list()` entries in `candidateModels()`.
```{r stvcLMstack_svc}
cand.svc <- candidateModels(list(phi_s = c(3, 6, 10), phi_t = 1,
                                 noise_sp_ratio = c(0.1, 0.5)), "cartesian")

mod5 <- stvcLMstack(y_svc ~ x1 + x2 + (x1), data = dat,
                    sp_coords = as.matrix(dat[, c("s1", "s2")]),
                    time_coords = matrix(0, n, 1),
                    cor.fn = "gneiting-decay",
                    process.type = "independent.shared",
                    candidate.models = cand.svc,
                    n.samples = 1000, verbose = TRUE)
post_svc <- stackedSampler(mod5)
```

The samples of $z$ are stacked by process: rows $1, \ldots, n$ hold the varying intercept and rows $n+1, \ldots, 2n$ the varying part of the slope of `x1`. The slope of `x1` at each location is $\beta_1 + z_2(s)$, whose true value is $5 + z(s)$.
```{r fig.align='center', fig.alt="Stacked posterior of the spatially varying slope against its true value"}
slope <- sweep(post_svc$z[n + seq_len(n), ], 2, post_svc$beta[2, ], "+")
slope_summ <- t(apply(slope, 1, function(x) quantile(x, c(0.025, 0.5, 0.975))))
slope_df <- data.frame(truth = 5 + dat$z_true, sL = slope_summ[, 1],
                       sM = slope_summ[, 2], sU = slope_summ[, 3])

ggplot(data = slope_df, aes(x = truth)) +
  geom_point(aes(y = sM), size = 0.75, color = "darkblue", alpha = 0.5) +
  geom_errorbar(aes(ymin = sL, ymax = sU), width = 0.05, alpha = 0.15,
                color = "skyblue") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  xlab("True slope") + ylab("Stacked posterior of slope") + theme_bw() +
  theme(panel.background = element_blank(),
        panel.grid = element_blank(), aspect.ratio = 1)
```

Finally, we compare the interpolated surfaces of the true slope and its posterior median.
```{r fig.align='center', fig.height=3.5, fig.width=7, fig.alt="Interpolated surfaces of the true spatially varying slope and its posterior median."}
dat$slope_true <- 5 + dat$z_true
dat$slope_hat <- slope_summ[, 2]
plot_slope <- surfaceplot2(dat, coords_name = c("s1", "s2"),
                           var1_name = "slope_true", var2_name = "slope_hat")
patchwork::wrap_plots(plot_slope) +
    patchwork::plot_layout(guides = "collect") &
    ggplot2::theme(legend.position = "right")
```

### Analysis of spatial non-Gaussian data

We also offer functions for Bayesian analysis of spatially point-referenced Poisson, binomial count, and binary data.

#### Spatial Poisson count data
We plot the Poisson counts `y_pois` at the same 200 locations.
```{r fig.align='center'}
ggplot(dat, aes(x = s1, y = s2)) +
  geom_point(aes(color = y_pois), alpha = 0.75) +
  scale_color_distiller(palette = "RdBu", direction = -1,
                        label = function(x) sprintf("%.0f", x)) +
  guides(alpha = 'none') + theme_bw() +
  theme(axis.ticks = element_line(linewidth = 0.25),
        panel.background = element_blank(), panel.grid = element_blank(),
        legend.title = element_text(size = 10, hjust = 0.25),
        legend.box.just = "center", aspect.ratio = 1)
```

#### Under fixed hyperparameters
Next, we demonstrate the function `spGLMexact()` which delivers posterior samples of the fixed effects and the spatial random effects. The option `family` must be specified correctly while using this function. For instance, in the following example, the formula `y_pois ~ x1 + x2` and `family = "poisson"` corresponds to the spatial regression model $$y(s) \sim \mathsf{Poisson} (\lambda(s)), \quad \log \lambda(s) = \beta_0 + \beta_1 x_1(s) + \beta_2 x_2(s) + z(s)\;.$$

We provide fixed values of the spatial process parameters and a boundary adjustment parameter, given by the argument `boundary`, which if not supplied, defaults to 0.5. For details on the priors and its default value, see function documentation.

```{r spGLMexact_Pois}
mod1 <- spGLMexact(y_pois ~ x1 + x2, data = dat, family = "poisson",
                   coords = as.matrix(dat[, c("s1", "s2")]), cor.fn = "matern",
                   spParams = list(phi = phi0, nu = nu0),
                   priors = list(nu.beta = 5, nu.z = 5),
                   boundary = 0.5,
                   n.samples = 1000, verbose = TRUE)
```

We next collect the samples of the fixed effects and summarize them. The true value of the fixed effects with which the data was simulated is $\beta = (2, -0.5, 0.3)$ (for more details, see the documentation of the data `simSpatial`).
```{r}
post_beta <- mod1$samples$beta
summary_beta <- t(apply(post_beta, 1, function(x) quantile(x, c(0.025, 0.5, 0.975))))
rownames(summary_beta) <- mod1$X.names
print(summary_beta)
```

#### Posterior recovery of scale parameters
The analytic tractability of the posterior distribution under the $\GCM$ framework is enabled by marginalizing out the scale parameters $\sigma^2_\beta$ and $\sigma^2_z$ associated with the fixed effects $\beta$ and the spatial random effects $z$, respectively. However, posterior samples of $\sigma^2_\beta$ and $\sigma^2_z$ can be recovered using the function `recoverGLMscale()`.
```{r}
mod1 <- recoverGLMscale(mod1)
```
We visualize the posterior distributions of $\sigma_\beta$ and $\sigma_z$ through histograms.
```{r fig.align='center', fig.height=3.5, fig.width=7, fig.alt="Posterior distributions of the scale parameters of the fixed and the random effects."}
post_scale_df <- data.frame(value = sqrt(c(mod1$samples$sigmasq.beta, mod1$samples$sigmasq.z)),
                            group = factor(rep(c("sigma.beta", "sigma.z"),
                                    each = length(mod1$samples$sigmasq.beta))))
ggplot(post_scale_df, aes(x = value)) +
  geom_density(fill = "lightblue", alpha = 0.6) +
  facet_wrap(~ group, scales = "free") + labs(x = "", y = "Density") +
  theme_bw() + theme(panel.background = element_blank(),
                     panel.grid = element_blank(), aspect.ratio = 1)
```

#### Using predictive stacking
Next, we move on to the function `spGLMstack()` that will implement our proposed stacking algorithm. The argument `loopd.controls` is used to provide details on what algorithm to be used to find LOO-PD. Valid options for the tag `method` is `"exact"` and `"CV"`. We use $K$-fold cross-validation by assigning `method = "CV"`and `CV.K = 10`. The tag `nMC` decides the number of Monte Carlo samples to be used to find the LOO-PD.
```{r spGLMstack_Pois}
cand.mod <- candidateModels(list(phi = c(3, 6, 10), nu = c(0.5, 1.5),
                                 boundary = c(0.5, 0.6)), "cartesian")

mod2 <- spGLMstack(y_pois ~ x1 + x2, data = dat, family = "poisson",
                   coords = as.matrix(dat[, c("s1", "s2")]), cor.fn = "matern",
                   candidate.models = cand.mod,
                   n.samples = 1000, priors = list(mu.beta = 5, nu.z = 5),
                   loopd.controls = list(method = "CV", CV.K = 10, nMC = 1000),
                   parallel = FALSE, verbose = TRUE)
```

We can extract information on the solver used for the stacking weights, its status, and the runtime by the following.
```{r}
print(mod2$diagnostics$solver$used)
print(mod2$diagnostics$solver$status)
print(mod2$run.time)
```

Further, we can recover the posterior samples of the scale parameters by passing the output obtained by running `spGLMstack()` once again through `recoverGLMscale()`.
```{r}
mod2 <- recoverGLMscale(mod2)
```

#### Sampling from stacked posterior
We first obtain final posterior samples by sampling from the stacked sampler.
```{r}
post_samps <- stackedSampler(mod2)
```

Subsequently, we summarize the posterior samples of the fixed effects.
```{r}
post_beta <- post_samps$beta
summary_beta <- t(apply(post_beta, 1, function(x) quantile(x, c(0.025, 0.5, 0.975))))
rownames(summary_beta) <- mod2$X.names
print(summary_beta)
```

The response `y_pois` was simulated using $\beta = (2, -0.5, 0.3)^\T$.
```{r, fig.align='center', fig.height=3, fig.width=8, fig.alt="Posterior distributions of the fixed effects"}
post_beta_df <- as.data.frame(post_beta)
post_beta_df <- post_beta_df %>%
  mutate(row = paste0("beta", row_number()-1)) %>%
  pivot_longer(-row, names_to = "sample", values_to = "value")

# True values of beta0, beta1 and beta2
truth <- data.frame(row = c("beta0", "beta1", "beta2"), true_value = c(2, -0.5, 0.3))

ggplot(post_beta_df, aes(x = value)) +
  geom_density(fill = "lightblue", alpha = 0.6) +
  geom_vline(data = truth, aes(xintercept = true_value),
             color = "red", linetype = "dashed", linewidth = 0.5) +
  facet_wrap(~ row, scales = "free") + labs(x = "", y = "Density") +
  theme_bw() + theme(panel.background = element_blank(),
                     panel.grid = element_blank(), aspect.ratio = 1)
```

Finally, we analyze the posterior samples of the spatial random effects.
```{r fig.align='center'}
post_z <- post_samps$z
post_z_summ <- t(apply(post_z, 1, function(x) quantile(x, c(0.025, 0.5, 0.975))))
z_combn <- data.frame(z = dat$z_true, zL = post_z_summ[, 1],
                      zM = post_z_summ[, 2], zU = post_z_summ[, 3])

plotz <- ggplot(data = z_combn, aes(x = z)) +
  geom_point(aes(y = zM), size = 0.75, color = "darkblue", alpha = 0.5) +
  geom_errorbar(aes(ymin = zL, ymax = zU), width = 0.05, alpha = 0.15,
                color = "skyblue") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  xlab("True z") + ylab("Stacked posterior of z") + theme_bw() +
  theme(panel.background = element_blank(),
        panel.grid = element_blank(), aspect.ratio = 1)
plotz
```

We can also compare the interpolated spatial surfaces of the true spatial effects with that of their posterior median.
```{r fig.align='center', fig.height=3.5, fig.width=7}
postmedian_z <- apply(post_z, 1, median)
dat$z_hat <- postmedian_z
plot_z <- surfaceplot2(dat, coords_name = c("s1", "s2"),
                       var1_name = "z_true", var2_name = "z_hat")
patchwork::wrap_plots(plot_z) +
    patchwork::plot_layout(guides = "collect") &
    ggplot2::theme(legend.position = "right")
```

### Spatial binomial count data
This will follow the same workflow as Poisson data with the exception that the structure of `formula` that defines the model will also contain the total number of trials at each location. We use the binomial counts `y_binom` out of `n_trials` trials at the same 200 locations. Here, we present only the `spGLMexact()` function for brevity.
```{r}
mod1 <- spGLMexact(cbind(y_binom, n_trials) ~ x1 + x2, data = dat, family = "binomial",
                   coords = as.matrix(dat[, c("s1", "s2")]), cor.fn = "matern",
                   spParams = list(phi = 6, nu = 0.5),
                   boundary = 0.5, n.samples = 1000, verbose = FALSE)
```

Similarly, we collect the posterior samples of the fixed effects and summarize them. The true value of the fixed effects with which the data was simulated is $\beta = (0.5, -0.5, 0.5)$.
```{r}
post_beta <- mod1$samples$beta
summary_beta <- t(apply(post_beta, 1, function(x) quantile(x, c(0.025, 0.5, 0.975))))
rownames(summary_beta) <- mod1$X.names
print(summary_beta)
```

### Spatial binary data
Finally, we present only the `spGLMexact()` function for the spatial binary data `y_bin` to avoid repetition. In this case, unlike the binomial model, almost nothing changes from that of in the case of spatial Poisson data.
```{r}
mod1 <- spGLMexact(y_bin ~ x1 + x2, data = dat, family = "binary",
                   coords = as.matrix(dat[, c("s1", "s2")]), cor.fn = "matern",
                   spParams = list(phi = 6, nu = 0.5),
                   boundary = 0.5, n.samples = 1000, verbose = FALSE)
```

Similarly, we collect the posterior samples of the fixed effects and summarize them. The true value of the fixed effects with which the data was simulated is $\beta = (0.5, -0.5, 0.5)$.
```{r}
post_beta <- mod1$samples$beta
summary_beta <- t(apply(post_beta, 1, function(x) quantile(x, c(0.025, 0.5, 0.975))))
rownames(summary_beta) <- mod1$X.names
print(summary_beta)
```

## References {-}

<div id="refs"></div>
